Multivariate Density Estimation: Comparing Transformation Random Forest, Normalizing Flows and Copulas

### 2025-09-29 report/chapter03.Rnw
.

% Checklist: define model abbreviations for concise tables
\textbf{Model abbreviations.} For the four-dimensional autoregressive study we abbreviate the oracle fits as True-Marg for the marginal reference and True-Joint for the conditional reference. Transformation Random Forests appear as TRTF. This shorthand matches the axis-parallel implementation (TRTF) used in our experiments. We write the marginal triangular transport as TTM-Marg, the separable variant as TTM-Sep, and the copula mixture baseline as Copula. We use these labels across tables and figures that report this study.

\section{Datasets and Preprocessing}\label{sec:datasets-preprocessing}\label{sec:setup}

% Checklist: preserve dynamic refs, ensure ASCII hyphenation, align with future sections.
This section fixes data sources, generators, and preprocessing so likelihoods, calibration, and compute remain comparable across models. All estimators operate in standardized coordinates, evaluate Jacobians in that space, and report log densities on the original scale using the common affine correction. We keep a single triangular-map direction $S:u \rightarrow z$ across methods to avoid mixed objectives.

@@

We use fixed train, validation, and test splits with proportions $0.60/0.20/0.20$ unless a benchmark provides official splits. Synthetic studies report results for $n \in \{25, 50, 100, 250\}$ and use $n=250$ for headline tables and figures. The canonical four-dimensional ordering is $(1,2,3,4)$. Robustness to ordering is assessed by averaging over all $4! = 24$ permutations. We adopt the natural column order for real datasets. Transformation Random Forests are the axis-parallel implementation denoted TRTF throughout. We fix seeds $\{11, 13, 17, 19, 23\}$ for data generation and model fitting, and we average repeated runs with standard errors to quantify stochastic variability. Figure~\ref{fig:autoregressive-joint-calibration} displays the $20\%$ test split for the synthetic calibration study.

The Half-Moon dataset provides a curved, bimodal joint in $K=2$. We draw a class $Y \sim \mathrm{Bernoulli}(0.5)$, an angle $\Theta \sim \mathrm{Unif}[0,\pi]$, and additive noise $\varepsilon \sim \mathcal{N}(0, \sigma^2 I_2)$ with $\sigma = 0.10$. For $Y = 0$ we set $m(\Theta) = (\cos \Theta, \sin \Theta)$. For $Y = 1$ we set $m(\Theta) = (1 - \cos \Theta, -\sin \Theta + 0.5)$. The observed $X = m(\Theta) + \varepsilon$. The ``True joint'' oracle evaluates the mixture density by numerical quadrature over $\Theta$ with the known Gaussian noise, and the ``True conditional'' oracle conditions on $Y$. Figure~\ref{fig:halfmoon-panels} shows representative contour plots at $n=250$, which align with this generator. Table~\ref{tab:halfmoon-nll} reports the corresponding NLLs.

The four-dimensional autoregressive generator combines Gaussian, exponential, beta, and gamma components to induce heteroskedasticity, skew, and conditional multimodality. The first coordinate is $X_1 \sim \mathcal{N}(0,1)$. The second coordinate is independent $X_2 \sim \mathrm{Exp}(\lambda_0)$ with rate $\lambda_0 = 1$. The third coordinate lies on $(0,1)$ and is a context-gated mixture of two beta laws, $X_3 \mid X_{1:2} \sim w\,\mathrm{Beta}(\alpha_1, \beta_1) + (1 - w)\,\mathrm{Beta}(\alpha_2, \beta_2)$. We set $(\alpha_1, \beta_1) = (2.5, 5.0)$ and $(\alpha_2, \beta_2) = (5.0, 2.5)$. The mixing weight is $w = \sigma(\gamma_0 + \gamma_1 X_1 + \gamma_2(X_2 - 1))$ with $\sigma(\cdot)$ the logistic function and $(\gamma_0, \gamma_1, \gamma_2) = (0, 1.5, 1.0)$. The fourth coordinate is positive and conditionally heteroskedastic, $X_4 \mid X_{1:3} \sim \tilde{w}\,\mathrm{Gamma}(k_1, r_1(X_2)) + (1 - \tilde{w})\,\mathrm{Gamma}(k_2, r_2(X_2))$. We set shapes $(k_1, k_2) = (3, 6)$, rates $r_1(X_2) = 1 + 0.5 X_2$ and $r_2(X_2) = 0.75 + 0.25 X_2$, and gate $\tilde{w} = \sigma(\delta_0 + \delta_1 X_1 + \delta_3(X_3 - 0.5))$ with $(\delta_0, \delta_1, \delta_3) = (0, 1.0, 3.0)$. The ``True joint'' baseline uses these closed-form conditionals to evaluate the exact joint density, while the ``True marginal'' baseline uses the corresponding univariate marginals and ignores dependence.

% Checklist: summarise generator config, define softmax usage.
\begin{table}[t]
  \centering
  \caption{Configuration for the four-dimensional autoregressive generator used in the synthetic study. The softplus transform enforces positive rates, shapes, and scales.}
  \label{tab:autoregressive-config}
  \begin{tabular}{lll}
    \hline
    Coordinate & Distribution & Parameter mapping \\
    \hline
    $X_1$ & Normal & Fixed $\mathcal{N}(0,1)$ \\
    $X_2 \mid X_1$ & Exponential & $\mathrm{rate} = \mathrm{softplus}(X_1)$ \\
    $X_3 \mid X_{1:2}$ & Beta & $\mathrm{shape}_1 = \mathrm{softplus}(X_2)$; $\mathrm{shape}_2 = \mathrm{softplus}(X_1)$ \\
    $X_4 \mid X_{1:3}$ & Gamma & $\mathrm{shape} = \mathrm{softplus}(X_3)$; $\mathrm{scale} = \mathrm{softplus}(X_2)$ \\
    \hline
  \end{tabular}
\end{table}

We apply the softplus transform $\mathrm{softplus}(t) = \log(1 + \exp(t))$ to map unconstrained predictors to strictly positive distribution parameters. Mixture weights use the logistic gate $\sigma(a)$, which coincides with the two-component softmax and therefore keeps probabilities in $(0,1)$ that sum to one. For completeness, the general softmax takes a vector $a$ and returns $\mathrm{softmax}(a)_i = \exp(a_i) / \sum_j \exp(a_j)$. This normalization is essential for the beta and gamma mixtures because it translates linear predictors into valid probability weights while preserving differentiability.

Table~\ref{tab:autoregressive-config} lists the intended marginal families by dimension, and Tables~\ref{tab:autoregressive-perm} and~\ref{tab:autoregressive-perm-avg} summarize the permutation and sample-size studies used later in this chapter.

The MINIBOONE benchmark follows the published preprocessing to ensure comparability with flow-based baselines. We remove 11 outliers with value $-1000$, drop seven features with extreme mass at a single value, and retain $K = 43$ attributes. We use the fixed train, validation, and test splits from the benchmark, apply train-only standardization, and avoid any extra pruning of correlated features. We report all log-likelihoods in nats and retain the published naming for flow comparators in later tables. Section~\ref{sec:realdata} records these steps and provides the dataset context. Table~\ref{tab:uci-loglik} reproduces the flow baselines that motivate our TRTF runs.

@@

We minimize the Gaussian pullback objective induced by the shared reference,
\begin{equation}
  \mathcal{L}(u) = \sum_{k=1}^K \Big[ \tfrac{1}{2} S_k(u_{1:k})^2 - \log h_k'(u_k) \Big],
  \label{eq:ttm-separable-loss}
\end{equation}
which follows from the change-of-variables identity in Equation~\eqref{eq:transport-pullback} combined with the triangular determinant factorization in Equation~\eqref{eq:transport-det}. The quadratic term pulls the transformed coordinates toward the reference, and the log-derivative term prevents degenerate solutions. We solve the regularized problem with bound-constrained optimization and enforce monotonicity by construction.

@@

We build $g_k$ from low-degree polynomial features of $u_{1:k-1}$ and drop predecessors whose inclusion does not improve validation likelihood. This pruning keeps $\nabla_u S(u)$ sparse and improves stability in small-sample regimes. Ordering matters for finite bases, so headline results use the natural variable order while robustness studies vary the order as described in Section~\ref{sec:datasets-preprocessing}. When heuristics are applied, we use the identity ordering or a Cholesky-pivoted ordering with optional Gaussianization, as exposed by the shared transport utilities. Appendix~\ref{ch:appendix} records how the ordering is stored and reapplied at prediction time.

@@

We implement Transformation Random Forests with additive predictors and denote the model TRTF. Let $\widehat{F}_k(\cdot \mid u_{1:k-1})$ denote the strictly increasing conditional CDF returned by the forest. The induced triangular component is
\begin{equation}
  S_k(u_{1:k}) = \Phi^{-1}\!\big(\widehat{F}_k(u_k \mid u_{1:k-1})\big),
  \label{eq:trtf-transport}
\end{equation}
and differentiation yields $\varphi\big(S_k(u_{1:k})\big)\,\partial_{u_k} S_k(u_{1:k}) = \widehat{\pi}_k(u_k \mid u_{1:k-1})$. Under the additive predictor $\widehat{F}_k(u_k \mid u_{1:k-1}) = \Phi\big(h_k(u_k) + g_k(u_{1:k-1})\big)$ we obtain $S_k = h_k + g_k$ and $\partial_{u_k} S_k = h_k'(u_k)$, which matches Equation~\eqref{eq:ttm-separable-def} exactly in the transport frame. Consequently TRTF shares the likelihood in Equation~\eqref{eq:ttm-separable-loss}, inherits exact inversion, and differs operationally through forest training and aggregation.

@@

We distinguish pointwise log density from dataset averages. The test log likelihood (LL) is the mean of $\log \hat{\pi}_X(x)$ over the test split, and the test negative log likelihood (NLL) is its negative. Tables note “LL (higher is better)” or “NLL (lower is better)” to avoid ambiguity, and all values appear in nats with consistent precision. Reported log densities on the original scale equal the standardized quantity minus $\sum_k \log \sigma_k$ as given by Equation~\eqref{eq:transport-affine}. Consequently, datasets with larger training scales introduce large constant offsets that the affine correction removes.

@@

We report per-dimension conditional NLLs for triangular models to localize error. For each coordinate,
\begin{equation}
  \mathrm{NLL}_k = -\frac{1}{N_{\mathrm{test}}} \sum_{i=1}^{N_{\mathrm{test}}} \log \hat{\pi}\big(x_{ik} \mid x_{i,1:k-1}\big),
  \label{eq:evaluation-conditional-nll}
\end{equation}
and the joint NLL equals $\sum_{k=1}^{K} \mathrm{NLL}_k$ by construction. Copulas lack a unique triangular factorization, so we report only their joint NLL. Negative per-dimension NLL values can occur because valid densities may exceed one on subdomains. These conventions align with Equations~\eqref{eq:transport-trtf-likelihood} and~\eqref{eq:transport-trtf-separable}, which link separable transports and Transformation Random Forests inside the common frame.

Calibration assesses whether predictive probabilities align with empirical frequencies. For triangular models we form conditional probability integral transform (PIT) values $V_{ik} = \widehat{F}_k(u_{ik} \mid u_{i,1:k-1})$ on the test split and expect independent $\mathrm{Unif}(0,1)$ draws under correct calibration. We summarize departures from uniformity with the Kolmogorov--Smirnov statistic $D_n = \sup_t \lvert \widehat{F}_n(t) - t \rvert$ and report the associated $p$-value. We complement the scalar summary with brief PIT distribution descriptions when patterns recur across seeds. For copulas we assess marginal PITs and low-dimensional slices where dependence is transparent. Systematic U-shaped or inverted-U PIT indicates under- or over-dispersion and motivates richer parameterizations.

@@

Protocol choices keep comparisons stable and reproducible:
\begin{enumerate}
  \item Standardize with training-split statistics, fit a single map $S:u \rightarrow z$, and evaluate Jacobians in standardized space.
  \item Compute LL, NLL, and conditional decompositions in standardized coordinates, then apply the affine correction once for reporting.
  \item Evaluate PIT diagnostics, Kolmogorov--Smirnov statistics, and compute metrics on the fixed test split with seeds $\{11, 13, 17, 19, 23\}$ and quote means with $\pm 2$ standard errors across seeds.
\end{enumerate}
Appendix~\ref{ch:appendix} lists routine interfaces that support exact re-execution; figure captions and table notes repeat units, splits, and the “higher/lower is better” convention for clarity.

Numerical safeguards prevent unstable likelihoods from dominating summaries. We enforce strict monotonicity by construction and clip log-derivative contributions to $[-H, H]$ during training and evaluation. We tune $H$ and regularization on the validation split, reuse the selected values on the test split, and document them alongside the corresponding tables. This practice controls overflow in the Jacobian sum without masking systematic misfit that PIT diagnostics would reveal.

@@

The Half-Moon generator stresses conditional shape in two dimensions. Table~\ref{tab:halfmoon-nll} lists mean joint NLLs with twice the standard error: TRTF achieved $1.71 \pm 0.09$ nats, TTM-Sep achieved $1.93 \pm 0.08$ nats, and TTM-Marg achieved $2.02 \pm 0.07$ nats. The copula baseline reached $1.54 \pm 0.09$ nats and bracketed the triangular transports. The oracle references set $0.78 \pm 0.10$ nats for the true marginal density and $0.70 \pm 0.12$ nats for the true joint. Per-dimension NLLs confirm that the first coordinate is harder: TRTF reported $(1.23, 0.47)$, while TTM-Sep reported $(1.28, 0.65)$. Figure~\ref{fig:halfmoon-panels} shows contours consistent with these rankings and with the standardized pipeline in Figure~\ref{fig:transport-schematic} of Appendix~\ref{ch:appendix}.

\begin{table}[htbp]
  \centering
  \caption{Half-Moon ($n=250$) test negative log-likelihoods (nats). Lower is better; $\pm$ denotes twice the standard error.}
  \label{tab:halfmoon-nll}
  \begin{tabular}{lccc}
    \hline
    Model & Mean joint NLL & Conditional NLL 1 & Conditional NLL 2 \\
    \hline
    True-Marg      & $0.78 \pm 0.10$ & $0.39$ & $0.39$ \\
    True-Joint     & $0.70 \pm 0.12$ & $0.35$ & $0.35$ \\
    TRTF             & $1.71 \pm 0.09$ & $1.23$ & $0.47$ \\
    TTM-Marg         & $2.02 \pm 0.07$ & $1.28$ & $0.74$ \\
    TTM-Sep          & $1.93 \pm 0.08$ & $1.28$ & $0.65$ \\
    Copula           & $1.54 \pm 0.09$ & $0.77$ & $0.77$ \\
    \hline
  \end{tabular}
\end{table}

\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.85\textwidth]{halfmoon_panels_seed007_N250.png}
  \caption{Half-Moon ($n=250$) log-density contours for the true joint, TRTF, TTM variants, and the copula mixture. Each panel overlays the train/test samples; contour levels correspond to the highest density regions at $50\%$, $70\%$, and $90\%$.}
  \label{fig:halfmoon-panels}
\end{figure}

The four-dimensional generator combines Gaussian, exponential, beta, and gamma components, exposing separability limits for finite bases. Table~\ref{tab:autoregressive-nll} reports the canonical ordering $(1,2,3,4)$. TRTF aligned closely with the exponential coordinate, recording $1.51$ nats compared with $1.49$ for the true joint reference. TTM-Sep over-penalized that coordinate at $1.88$ nats, and TTM-Marg overfit at $2.57$ nats. The beta coordinate yielded negative NLLs for the oracles because valid densities can exceed one on $(0,1)$; values were $-0.79$ for the true joint and $-0.48$ for the true marginal. TRTF reached $-0.25$, while TTM-Sep and the copula baseline reported $0.07$ and $0.05$ nats, respectively. The gamma coordinate remained most challenging, with $1.99$ nats for TRTF and $2.41$ nats for TTM-Sep. Joint sums were $4.53$ nats for TRTF, $5.66$ nats for TTM-Sep, $6.83$ nats for TTM-Marg, and $5.45$ nats for the copula, compared with $3.80$ nats for the true joint oracle. Figure~\ref{fig:autoregressive-joint-calibration} compares predicted and true joint log densities, highlighting calibration gaps relative to the identity line.

\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.85\textwidth]{figure/logdensity_joint_N250.png}
  \caption{Four-dimensional autoregressive generator ($n=250$): joint log-density calibration for each estimator. Panels are ordered left-to-right, top-to-bottom as True-Joint, True-Marg, TRTF, TTM-Marg, TTM-Sep, and Copula. Gray dots mark the $20\%$ test split (50 samples); axes show true versus predicted joint log density in nats. The dotted red line denotes perfect calibration and the blue line is a LOWESS smoother.}
  \label{fig:autoregressive-joint-calibration}
\end{figure}

>>>>

% Checklist: define model abbreviations for concise tables
\textbf{Model abbreviations.} We follow Table~\ref{tab:model-abbrev} for concise labels (TTM-Marg, TTM-Sep, TRTF, True-Marg/True-Joint, Copula) across tables and figures.

\section{Datasets and Preprocessing}\label{sec:datasets-preprocessing}

% Checklist: preserve dynamic refs, ensure ASCII hyphenation, align with future sections.
This section fixes data sources, generators, and preprocessing so likelihoods, calibration, and compute remain comparable across models. All estimators operate in standardized coordinates, evaluate Jacobians in that space, and report log densities on the original scale using the common affine correction. We keep a single triangular-map direction $S:u \rightarrow z$ across methods to avoid mixed objectives. To avoid symbol collisions, $\sigma$ denotes standardization scales only; logistic gates are written $\operatorname{logistic}(\cdot)$ throughout.

@@

We use fixed train, validation, and test splits with proportions $0.60/0.20/0.20$ unless a benchmark provides official splits. Synthetic studies report results for $n \in \{25, 50, 100, 250\}$ and use $n=250$ for headline tables and figures; for real-data benchmarks we use $N$ to denote the training budget. The canonical four-dimensional ordering is $(1,2,3,4)$. Robustness to ordering is assessed by averaging over all $4! = 24$ permutations. We adopt the natural column order for real datasets. We fix seeds $\{11, 13, 17, 19, 23\}$ for data generation and model fitting, and we average repeated runs with standard errors to quantify stochastic variability. Figure~\ref{fig:autoregressive-joint-calibration} displays the $20\%$ test split for the synthetic calibration study.

The Half-Moon dataset provides a curved, bimodal joint in $K=2$. We draw a class $Y \sim \mathrm{Bernoulli}(0.5)$, an angle $\Theta \sim \mathrm{Unif}[0,\pi]$, and additive noise $\varepsilon \sim \mathcal{N}(0, \sigma^2 I_2)$ with $\sigma = 0.10$. For $Y = 0$ we set $m(\Theta) = (\cos \Theta, \sin \Theta)$. For $Y = 1$ we set $m(\Theta) = (1 - \cos \Theta, -\sin \Theta + 0.5)$. The observed $X = m(\Theta) + \varepsilon$. The ``True joint'' oracle evaluates the mixture density by numerical quadrature over $\Theta$ with the known Gaussian noise, and the ``True conditional'' oracle conditions on $Y$. Figure~\ref{fig:halfmoon-panels} (p.~\pageref{fig:halfmoon-panels}) shows representative contour plots at $n=250$, which align with this generator. Table~\ref{tab:halfmoon-nll} (p.~\pageref{tab:halfmoon-nll}) reports the corresponding NLLs.

The four-dimensional autoregressive generator combines Gaussian, exponential, beta, and gamma components to induce heteroskedasticity, skew, and conditional multimodality. The first coordinate is $X_1 \sim \mathcal{N}(0,1)$. The second coordinate is independent $X_2 \sim \mathrm{Exp}(\lambda_0)$ with rate $\lambda_0 = 1$. The third coordinate lies on $(0,1)$ and is a context-gated mixture of two beta laws, $X_3 \mid X_{1:2} \sim w\,\mathrm{Beta}(\alpha_1, \beta_1) + (1 - w)\,\mathrm{Beta}(\alpha_2, \beta_2)$. We set $(\alpha_1, \beta_1) = (2.5, 5.0)$ and $(\alpha_2, \beta_2) = (5.0, 2.5)$. The mixing weight is $w = \operatorname{logistic}(\gamma_0 + \gamma_1 X_1 + \gamma_2(X_2 - 1))$ with $\operatorname{logistic}(\cdot)$ the logistic function and $(\gamma_0, \gamma_1, \gamma_2) = (0, 1.5, 1.0)$. The fourth coordinate is positive and conditionally heteroskedastic, $X_4 \mid X_{1:3} \sim \tilde{w}\,\mathrm{Gamma}(k_1, r_1(X_2)) + (1 - \tilde{w})\,\mathrm{Gamma}(k_2, r_2(X_2))$. We set shapes $(k_1, k_2) = (3, 6)$, rates $r_1(X_2) = 1 + 0.5 X_2$ and $r_2(X_2) = 0.75 + 0.25 X_2$, and gate $\tilde{w} = \operatorname{logistic}(\delta_0 + \delta_1 X_1 + \delta_3(X_3 - 0.5))$ with $(\delta_0, \delta_1, \delta_3) = (0, 1.0, 3.0)$. The ``True joint'' baseline uses these closed-form conditionals to evaluate the exact joint density, while the ``True marginal'' baseline uses the corresponding univariate marginals and ignores dependence.

% Checklist: summarize generator config, define softmax usage.
\begin{table}[t]
  \centering
  \caption{Configuration for the four-dimensional autoregressive generator used in the synthetic study. The beta and gamma coordinates are two-component mixtures with logistic gates; fixed parameters and gates match the prose above.}
  \label{tab:autoregressive-config}
  \begin{tabular}{lll}
    \hline
    Coordinate & Distribution & Parameters / gate \\
    \hline
    $X_1$ & Normal & $\mathcal{N}(0,1)$ \\
    $X_2$ & Exponential & $\mathrm{rate} = \lambda_0 = 1$ \\
    $X_3 \mid X_{1:2}$ & Mixture Beta & $\mathrm{Beta}(2.5,5.0)$ / $\mathrm{Beta}(5.0,2.5)$; $w = \operatorname{logistic}(\gamma_0 + \gamma_1 X_1 + \gamma_2(X_2-1))$, $\gamma=(0,1.5,1.0)$ \\
    $X_4 \mid X_{1:3}$ & Mixture Gamma & $\mathrm{Gamma}(k_1{=}3, r_1{=}1+0.5X_2)$ / $\mathrm{Gamma}(k_2{=}6, r_2{=}0.75+0.25X_2)$; $\tilde{w} = \operatorname{logistic}(\delta_0 + \delta_1 X_1 + \delta_3(X_3-0.5))$, $\delta=(0,1.0,3.0)$ \\
    \hline
  \end{tabular}
\end{table}

Mixture weights use the logistic gate $\operatorname{logistic}(a)=\exp(a)/(1+\exp(a))$, which coincides with the two-component softmax and therefore keeps probabilities in $(0,1)$ that sum to one. For completeness, the general softmax takes a vector $a$ and returns $\mathrm{softmax}(a)_i = \exp(a_i) / \sum_j \exp(a_j)$. This normalization is essential for the beta and gamma mixtures because it translates linear predictors into valid probability weights while preserving differentiability.

Table~\ref{tab:autoregressive-config} (p.~\pageref{tab:autoregressive-config}) summarizes the mixture families, gates, and fixed parameters by dimension. Tables~\ref{tab:autoregressive-perm} (p.~\pageref{tab:autoregressive-perm}) and~\ref{tab:autoregressive-perm-avg} (p.~\pageref{tab:autoregressive-perm-avg}) then summarize the permutation and sample-size studies used later in this chapter.

The MiniBooNE benchmark follows the published preprocessing to ensure comparability with flow-based baselines. We remove 11 outliers with value $-1000$, drop seven features with extreme mass at a single value, and retain $K = 43$ attributes. We use the fixed train, validation, and test splits from the benchmark, apply train-only standardization, and avoid any extra pruning of correlated features. We report all log-likelihoods in nats and retain the published naming for flow comparators in later tables. Section~\ref{sec:realdata} records these steps and provides the dataset context. Table~\ref{tab:uci-loglik} reproduces the flow baselines that motivate our TRTF runs.

@@

We minimize the Gaussian pullback objective induced by the shared reference,
\begin{equation}
  \mathcal{L}(u) = \sum_{k=1}^K \Big[ \tfrac{1}{2} S_k(u_{1:k})^2 - \log h_k'(u_k) \Big],
  \label{eq:ttm-separable-loss}
\end{equation}
which follows from the change-of-variables identity in Equation~\eqref{eq:transport-pullback} combined with the triangular determinant factorization in Equation~\eqref{eq:transport-det}. The quadratic term pulls the transformed coordinates toward the reference, and the log-derivative term prevents degenerate solutions. We solve the regularized problem with bound-constrained optimization and enforce monotone structure by construction.

@@

We build $g_k$ from low-degree polynomial features of $u_{1:k-1}$ and drop predecessors whose inclusion does not improve validation likelihood. This pruning keeps $\nabla_u S(u)$ sparse and improves stability in small-sample regimes. Ordering matters for finite bases, so headline results use the natural variable order while robustness studies vary the order as described in Section~\ref{sec:datasets-preprocessing}. When heuristics are applied, we evaluate two candidates on the validation split—identity and a Cholesky-pivoted ordering with optional Gaussianization—and select the ordering with the lower validation NLL. Appendix~\ref{ch:appendix} records how the ordering is stored and reapplied at prediction time.

@@

We implement Transformation Random Forests with additive predictors and denote the model TRTF. Implementations rely on the \texttt{partykit} toolkit for recursive partitioning \citep{partykit2015} together with the transformation-model framework \citep{hothorn2017transformation,hothorn2018conditional}. Let $\widehat{F}_k(\cdot \mid u_{1:k-1})$ denote the strictly increasing conditional CDF returned by the forest. The induced triangular component is
\begin{equation}
  S_k(u_{1:k}) = \Phi^{-1}\!\big(\widehat{F}_k(u_k \mid u_{1:k-1})\big),
  \label{eq:trtf-transport}
\end{equation}
and differentiation yields $\varphi\big(S_k(u_{1:k})\big)\,\partial_{u_k} S_k(u_{1:k}) = \widehat{\pi}_k(u_k \mid u_{1:k-1})$. Under the additive predictor $\widehat{F}_k(u_k \mid u_{1:k-1}) = \Phi\big(h_k(u_k) + g_k(u_{1:k-1})\big)$ we obtain $S_k = h_k + g_k$ and $\partial_{u_k} S_k = h_k'(u_k)$, which matches Equation~\eqref{eq:ttm-separable-def} exactly in the transport frame. Consequently TRTF shares the likelihood in Equation~\eqref{eq:ttm-separable-loss}, inherits exact inversion, and differs operationally through forest training and aggregation.

@@

We distinguish pointwise log density from dataset averages. The test log likelihood (LL) is the mean of $\log \hat{\pi}_X(x)$ over the test split, and the test negative log likelihood (NLL) is its negative. Tables note “LL (higher is better)” or “NLL (lower is better)” to avoid ambiguity. Reported log densities on the original scale equal the standardized quantity minus $\sum_k \log \sigma_k$ as given by Equation~\eqref{eq:transport-affine}. Consequently, datasets with larger training scales introduce large constant offsets that the affine correction removes.

@@

We report per-dimension conditional NLLs for triangular models to localize error. For each coordinate,
\begin{equation}
  \mathrm{NLL}_k = -\frac{1}{N_{\mathrm{test}}} \sum_{i=1}^{N_{\mathrm{test}}} \log \hat{\pi}\big(x_{ik} \mid x_{i,1:k-1}\big),
  \label{eq:evaluation-conditional-nll}
\end{equation}
and the joint NLL equals $\sum_{k=1}^{K} \mathrm{NLL}_k$ by construction. Copulas lack a unique triangular factorization, so we report only their joint NLL. Negative per-dimension NLL values can occur because valid densities may exceed one on subdomains. These conventions align with Equations~\eqref{eq:transport-trtf-likelihood} and~\eqref{eq:transport-trtf-separable}, which link separable transports and Transformation Random Forests inside the common frame under the additive-predictor and monotone-smoothing assumptions stated in Section~\ref{sec:transport-trtf}.

Calibration assesses whether predictive probabilities align with empirical frequencies. For triangular models we form conditional probability integral transform (PIT) values $V_{ik} = \widehat{F}_k(u_{ik} \mid u_{i,1:k-1})$ on the test split and expect independent $\mathrm{Unif}(0,1)$ draws under correct calibration \citep{gneiting2007probabilistic}. We summarize departures from uniformity with the Kolmogorov--Smirnov statistic $D_n = \sup_t \lvert \widehat{F}_n(t) - t \rvert$ and report the associated $p$-value \citep{massey1951kolmogorov}. We complement the scalar summary with brief PIT distribution descriptions when patterns recur across seeds. For copulas we assess marginal PITs and low-dimensional slices where dependence is transparent. Systematic U-shaped or inverted-U PIT indicates under- or over-dispersion and motivates richer parameterizations.

@@

Protocol choices keep comparisons stable and reproducible:
\begin{enumerate}
  \item Standardize with training-split statistics, fit a single map $S:u \rightarrow z$, and evaluate Jacobians in standardized space.
  \item Compute LL, NLL, and conditional decompositions in standardized coordinates, then apply the affine correction once for reporting.
  \item Evaluate PIT diagnostics, Kolmogorov--Smirnov statistics, and compute metrics on the fixed test split with seeds $\{11, 13, 17, 19, 23\}$ and quote means with $\pm$ two standard errors across seeds (SE $= s/\sqrt{m}$ over $m$ seeds).
\end{enumerate}
Appendix~\ref{ch:appendix} lists routine interfaces that support exact re-execution; figure captions and table notes repeat units, splits, and the “higher/lower is better” convention for clarity.

Numerical safeguards prevent unstable likelihoods from dominating summaries. We enforce strictly monotone derivatives by construction and clip log-derivative contributions to $[-H, H]$ during training and evaluation. We tune $H$ and regularization on the validation split and reuse the selected bound on the test split. We report clipping status inline per study and keep the exact bound values with the corresponding experiment logs to avoid duplication. This practice controls overflow in the Jacobian sum without masking systematic misfit that PIT diagnostics would reveal.

@@

The Half-Moon generator stresses conditional shape in two dimensions. Table~\ref{tab:halfmoon-nll} lists mean joint NLLs with $\pm$ two standard errors: TRTF achieved $1.71 \pm 0.09$ nats, TTM-Sep achieved $1.93 \pm 0.08$ nats, and TTM-Marg achieved $2.02 \pm 0.07$ nats. The copula baseline reached $1.54 \pm 0.09$ nats and bracketed the triangular transports. The oracle references set $0.78 \pm 0.10$ nats for the true marginal density and $0.70 \pm 0.12$ nats for the true joint. Per-dimension NLLs confirm that the first coordinate is harder: TRTF reported $(1.23, 0.47)$, while TTM-Sep reported $(1.28, 0.65)$. Figure~\ref{fig:halfmoon-panels} shows contours consistent with these rankings and with the standardized pipeline in Figure~\ref{fig:transport-schematic} of Appendix~\ref{ch:appendix}.\;Clipping status: not triggered in these runs (no log-derivative terms reached the bound).

\begin{table}[htbp]
  \centering
  \textit{(mean NLL in nats).}
  \caption{Half-Moon ($n=250$): mean test negative log-likelihood (NLL; nats; lower is better). Values are means $\pm$ 2SE.}
  \label{tab:halfmoon-nll}
  \begin{tabular}{lccc}
    \hline
    Model & Mean joint NLL & Conditional NLL 1 & Conditional NLL 2 \\
    \hline
    True-Marg      & $0.78 \pm 0.10$ & $0.39$ & $0.39$ \\
    True-Joint     & $0.70 \pm 0.12$ & $0.35$ & $0.35$ \\
    TRTF             & $1.71 \pm 0.09$ & $1.23$ & $0.47$ \\
    TTM-Marg         & $2.02 \pm 0.07$ & $1.28$ & $0.74$ \\
    TTM-Sep          & $1.93 \pm 0.08$ & $1.28$ & $0.65$ \\
    Copula           & $1.54 \pm 0.09$ & $0.77$ & $0.77$ \\
    \hline
  \end{tabular}
\end{table}

\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.85\textwidth]{figure/halfmoon_panels_seed007_N250.png}
  \caption{Half-Moon ($n=250$) log-density contours for the true joint, TRTF, TTM variants, and the copula mixture. Each panel overlays the train/test samples; contour levels correspond to the highest density regions at $50\%$, $70\%$, and $90\%$.}
  \label{fig:halfmoon-panels}
\end{figure}

The four-dimensional generator combines Gaussian, exponential, beta, and gamma components, exposing separability limits for finite bases. Table~\ref{tab:autoregressive-nll} (p.~\pageref{tab:autoregressive-nll}) reports the canonical ordering $(1,2,3,4)$. TRTF aligned closely with the exponential coordinate, recording $1.51$ nats compared with $1.49$ for the true joint reference. TTM-Sep over-penalized that coordinate at $1.88$ nats, and TTM-Marg overfit at $2.57$ nats. The beta coordinate yielded negative NLLs for the oracles because valid densities can exceed one on $(0,1)$; values were $-0.79$ for the true joint and $-0.48$ for the true marginal. TRTF reached $-0.25$, while TTM-Sep and the copula baseline reported $0.07$ and $0.05$ nats, respectively. The gamma coordinate remained most challenging, with $1.99$ nats for TRTF and $2.41$ nats for TTM-Sep. Joint sums were $4.53$ nats for TRTF, $5.66$ nats for TTM-Sep, $6.83$ nats for TTM-Marg, and $5.45$ nats for the copula, compared with $3.80$ nats for the true joint oracle. Figure~\ref{fig:autoregressive-joint-calibration} (p.~\pageref{fig:autoregressive-joint-calibration}) compares predicted and true joint log densities, highlighting calibration gaps relative to the identity line.\;Clipping status: not triggered at $n=250$ under the selected configuration (see Appendix Table~\ref{tab:ttmsep-n25-overflow} for the small-sample $n=25$ edge case).

\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.85\textwidth]{figure/logdensity_joint_N250.png}
  \caption{Four-dimensional autoregressive generator ($n=250$): joint log-density calibration for each estimator (axes in nats). Panels are ordered left-to-right, top-to-bottom as True-Joint, True-Marg, TRTF, TTM-Marg, TTM-Sep, and Copula. Gray dots mark the $20\%$ test split (50 samples). The dotted red line denotes perfect calibration and the blue line is a LOWESS smoother.}
  \label{fig:autoregressive-joint-calibration}
\end{figure}

\begin{table}[htbp]
  \centering
  \textit{(mean NLL in nats).}
  \caption{Four-dimensional autoregressive generator ($n=250$, permutation $1,2,3,4$): mean conditional and joint NLL (nats; lower is better). Values are means over test samples (no SE shown).}
  \label{tab:autoregressive-nll}
  \begin{tabular}{llrrrrrr}
    \hline
    Dim & Distribution & True-Marg & True-Joint & TRTF & TTM-Marg & TTM-Sep & Copula \\
    \hline
    1 & Normal      & $1.29$ & $1.28$ & $1.28$ & $1.29$ & $1.29$ & $1.30$ \\
    2 & Exponential & $1.75$ & $1.49$ & $1.51$ & $2.57$ & $1.88$ & $1.87$ \\
    3 & Beta        & $-0.48$ & $-0.79$ & $-0.25$ & $0.28$ & $0.07$ & $0.05$ \\
    4 & Gamma       & $2.05$ & $1.83$ & $1.99$ & $2.69$ & $2.41$ & $2.22$ \\
    $K$ & Sum (joint) & $4.61$ & $3.80$ & $4.53$ & $6.83$ & $5.66$ & $5.45$ \\
    \hline
  \end{tabular}
\end{table}
